Psychological Needs During Intergroup Contact

### 2022-01-17 manuscript.tex
Abstract and Significance Statement

% Abstract
\abstract{
Abstract goes here.

\noindent\textbf{Public significance statement}: Significance Statement goes here.
}

>>>>

% Abstract
\abstract{
One challenge of modern intergroup contact research has been a question of when and why an interaction is perceived as positive and carries the benefits of positive intergroup contact. The understudied mechanism of interaction quality is highlighted by a broad body of literature showing the adverse effects of negative intergroup interactions. To understand positive intergroup interactions, we propose to look at situational motivations and psychological needs in every-day intergroup interactions. We collected three extensive longitudinal studies of recent migrants to capture their interactions with the majority outgroup (total \textit{N} of measurements = 10,297). Across the three studies, we find inconsistent evidence for interaction frequencies and average interaction quality in predicting outgroup attitudes. However, using non-aggregated data shows a strong influence of outgroup interactions on outgroup attitudes. We find that the fulfillment of psychological needs offers a strong predictor of the positive attitudes after interactions and that these positive effects go through perceived interaction quality. The situational needs remained the core predictor of outgroup attitudes even after controlling for Allport's contact conditions or more general fundamental needs. As one of the first studies to test intergroup contact theory using extensive longitudinal data, we offer insight into the mechanisms of positive intergroup contact during real-life interactions and find situational motivations to be a key building block of understanding and addressing positive intergroup interactions.

\noindent\textbf{Public significance statement}: In this paper we provide evidence that the fulfillment of situational psychological needs during real-life intergroup contacts meaningfully predicts perceived interaction quality and positive outgroup attitudes. This offers testament both to emerging practice of capturing real-life interactions using extensive longitudinal data and that motivational fulfillment offer a flexible and relevant mechanism of understanding positive intergroup contact.
}
